```{r echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = "figures/")
options(scipen = 1)
```


# Reproducible Research: Peer Assessment 1

We'll need some extra packages for the analysis:

```{r}
library(ggplot2)
library(dplyr)
```


## Loading and preprocessing the data

First, we load the raw data and transform the date variable to date format.

```{r cache = TRUE}
activity <- read.csv(unz("activity.zip", "activity.csv"),
                     colClasses = c(NA, "Date", NA))
```


## What is mean total number of steps taken per day?

Now, we calculate the total number of steps taken each day and plot the distribution of the steps.

```{r cache = TRUE}
daily_steps <- activity %>%
        group_by(date) %>%
        select(steps, date) %>%
        summarise(
                  sum = sum(steps, na.rm = TRUE)
                 )

qplot(sum, 
      data = daily_steps, 
      binwidth = 1000,
      xlab = "Steps per day")
```

On the aggregated dataset we can calculate some descriptive statistics -- including the mean and median -- of the total number of steps taken per day.

```{r}
summary(daily_steps$sum)
```


## What is the average daily activity pattern?

```{r cache = TRUE}
daily_pattern <- activity %>%
        group_by(interval) %>%
        select(steps, interval) %>%
        summarise(
                  ave = mean(steps, na.rm = TRUE)
                 )

qplot(interval, ave, 
      data = daily_pattern, 
      geom = "line",
      xlab = "Interval",
      ylab = "Number of steps")
```

We can retrive which inteval contains the maximum value in the following way:

```{r}
daily_pattern[which.max(daily_pattern$ave), ]
```

On average across all the days the 5-minute interval when the maximum number of steps was `r daily_pattern[which.max(daily_pattern$ave), 1]` with `r daily_pattern[which.max(daily_pattern$ave), 2]` steps.

## Imputing missing values

The dataset contains many missing values:

```{r}
sum(is.na(activity$steps))
```

For every missing interval we impute the median value for that interval in a copy of the original dataset.

```{r}
activity_imputed <- activity %>%
        group_by(interval) %>%
        mutate(
                steps = ifelse(is.na(steps),
                               median(steps, na.rm = TRUE),
                               steps)
                )
```

Now, we calculate the number of steps per day in the imputed dataset and plot its distribution.

```{r}
daily_steps_imputed <- activity_imputed %>%
        group_by(date) %>%
        select(steps, date) %>%
        summarise(
                  sum = sum(steps)
                 )

qplot(sum, 
      data = daily_steps_imputed, 
      binwidth = 1000,
      xlab = "Steps per day")
```

In the imputed dataset the mean total number of steps taken per day was `r summary(daily_steps_imputed$sum)[4]` and the median was `r summary(daily_steps_imputed$sum)[3]`. As expected, the median didn't change while the mean increased slightly. As we compare the histograms of the original and imputed datasets, we can see this change: the distribution is less skewed to the left. Despite the imputation, there are still many days with very few steps, which means, the distribution is still bimodal.


## Are there differences in activity patterns between weekdays and weekends?

After some transformation we can plot the two series.

```{r cache = TRUE}
weekdays <- activity_imputed %>%
        mutate(
                weekday = weekdays(date) != "Saturday" &
                          weekdays(date) != "Sunday",
                weekday = factor(weekday,
                                 levels = c(FALSE, TRUE),
                                 labels = c("weekend", "weekday"))
                ) %>%
        group_by(interval, weekday) %>%
        select(steps, interval, weekday) %>%
        summarise(ave = mean(steps, na.rm = TRUE))

qplot(interval, ave, 
      data = weekdays,
      facets = weekday ~ ., 
      geom = "line",
      xlab = "Interval",
      ylab = "Number of steps")
```
